\chapter{Motivation}%
\label{ch:motivation}

Simplicity is a language by Russell O'Connor that has been in development since 2015.
It can be used to express spending conditions that lock coins (UTXOs) on a blockchain,
such as (multi)signatures,
hash- and timelocks,
covenants and much more.
It may be used as a replacement of Bitcoin Script in the future.
% TODO: reword
In this document,
we give an overview of Simplicity,
starting with the design goals behind it.

\section{Declarative}

% It is important to keep the naming consistent:
% Bitcoin Script has "scripts"
% Miniscript has "policies"
% Simplicity has "programs"

Expressing spending conditions in Bitcoin Script \emph{(here, we refer to this expression as \enquote{script})}
can be difficult without introducing errors.
In response,
Miniscript was developed as a language for expressing spending conditions in declarative spending policies
that let users focus on who can spend what.
On the blockchain,
the policy is converted into an equivalent script,
hiding the underlying implementation details of Bitcoin Script from the user.
Therefore,
Miniscript can be seen as a subset of Bitcoin Script with a human-friendly interface for expressing spending conditions.
%
Arguably,
this is a more natural approach.
It prevents bugs by making scripts more human-readable
and by making the semantics of a script---%
i.e., the conditions that it expresses---%
more explicit.
We can statically check properties such as
correctness \emph{(whether the resulting script can be redeemed on the blockchain without violating consensus rules)},
malleability \emph{(whether the script can be redeemed with a witness that does not satisfy all spending conditions)}
and fees \emph{(how large the resulting script plus witness is and how expensive this size is at the current rate)}.
%
Simplicity goes one step further: Simplicity is a declarative \emph{runtime} environment:
We remove the restriction of a stack,
inherent to Bitcoin Script,
and instead evaluate pure mathematical functions.
Spending conditions are expressed in \emph{programs} that are themselves functions,
built by combining a range of base functions as fundamental building blocks.
The Simplicity Bit Machine then executes these programs for us,
hiding the implementation details from the user.
As a result,
we get all the benefits of Miniscript and much more,
as we will see in the following sections.

% Idea: what a script expresses (i.e., a pure function over spending scenarios)
% and what a script _is_ (a DAG that is very close to a function) become one in Simplicity

\section{Verifiable}

The blockchain is irreversible,
so we want to absolutely make sure that our coins are locked correctly before we put them on there.
Simplicity is mathematically defined,
which means that the spending conditions expressed in a Simplicity program are clearly defined.
They are independent of any specific implementation in any given programming language.
We can use logical reasoners and theorem provers---%
such as the \href{https://coq.inria.fr/}{Coq proof assistant}---%
to prove,
for instance,
that a signature matching a given public key is the only valid witness for a program
and that any other witness is invalid.
We can also prove how many resources a program will use on the Bit Machine,
which implies fees.
Tasks like these are already hard to do for Miniscript and the underlying Bitcoin Script,
but Simplicity allows us to prove \emph{any} mathematical property of a given program.
If a property holds for a program,
then we can prove it.
The use of a verifiable language on the blockchain opens up entirely new possibilities for formal verification:
We can essentially \emph{prove} that our programs have no bugs.
By proving the security properties first,
we can avoid catastrophes like the Parity bug or DAO hack on Ethereum,
which were both caused by faulty scripts.
%
\enquote{Don't trust. Verify.}
Blockstream's company motto is stronger than ever in Simplicity.

\section{Expressive}

We want to be able to express as many spending conditions as possible.
Miniscript can only express what Bitcoin Script can express,
and Bitcoin Script cannot express everything,
so it is regularly extended with new op codes.
Simplicity can express \emph{every} clearly defined---%
i.e., decidable---%
spending condition out of the box.
This enables schemes like covenants, weighted thresholds and oracles,
and novel cryptography like zero-knowledge proofs and identity-based encryption.
The possible use cases that could build on this foundation are endless.
%
With great power comes great responsibility,
and with powerful scripts comes the risk of rogue scripts.
Simplicity provides all of its expressivity while being verifiable.
We can do great things, but we can also prove that we made no mistakes.
Figuratively speaking,
we tame the expressivity beast with the restraints of formal proof.
%
Also,
Simplicity is \emph{not} Turing complete,
which means that programs never run into infinite loops.
They finish after a short while,
whose duration we can computationally predict beforehand.

\section{Efficient}

It should be as fast as possible to evaluate
if a given witness satisfies the spending conditions of a given expression.
Simplicity's introspection is limited to the current transaction,
which means that programs can be evaluated on Bitcoin full nodes without the need for new index structures.
%
The program size is minimized for storage on the blockchain and for transmission:
Parts of a program that do the same computation are collapsed into one \emph{(called sharing)}.
Furthermore,
branches of a program that are not visited by a given witness are removed \emph{(called pruning)}.
The latter is similar to how Taproot hides unused scripts,
but Simplicity also hides unused branches in a \emph{used} program.
%
Expensive operations---%
such as signature checks---%
are implemented as \emph{jets}:
Special program expressions that call verified C code during their evaluation.
%
Finally,
being able to predict the resource usage of a program on the Bit Machine
enables runtime optimizations.

% TODO: summary of each chapter
